Inaction on climate change could approach one year of life in some European countries \thanks{All data and code that supports these conclusions are available as supplementary materials.}

### 2018-08-22 MS-cclifeexpec.tex
New branch

\title{Climate change projected to reduce European life expectancy \thanks{All data and code that supports these conclusions are available as
supplementary materials.}  }

@@

\usepackage{natbib}
\bibliographystyle{nature}
\usepackage[strings]{underscore} % protect underscores in most circumstances

@@

\@ifpackageloaded{color}{
    \PassOptionsToPackage{usenames,dvipsnames}{color}
}{%
    \usepackage[usenames,dvipsnames]{color}
}
\makeatother
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Mathew E. Hauer\textsuperscript{1}* (University of Georgia) and Alexis R. Santos\textsuperscript{2} (Pennsylvania State University)},
             pdfkeywords = {Climate change, Life expectancy, Mortality, Demography, Excess
mortality, Europe, Public health},  
            pdftitle={Climate change projected to reduce European life expectancy},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

\noindent Climate change related estimates of excess mortality clearly demonstrate
the dramatic impact on public health and human mortality from climate
change
\citep{patz2005impact, aastrom2013attributing, forzieri2017increasing, staddon2014climate}.
However, life expectancy at birth is more easily communicated and
understood by the public, and controls for different geographic,
economic, and demographic groups \citep{parrish2010peer}. No studies
have comprehensively connected excess mortality due to climate change to
anticipated reductions in life expectancy. Without properly situating
the potential loss of life within the contexts of life expectancy, we
risk misrepresenting cost of climate change on mortality. In this paper,
we convert excess mortality estimates into potential reductions of life
expectancy at birth and report the cost of climate change on human
longevity. We find climate change could become the third largest life
expectancy reducer behind heart disease and cancer in some European
countries. Thus, the cost of inaction on climate change could approach
one year of life in some countries.

@@

\textsuperscript{1} Carl Vinson Institute of Government, University of
Georgia. 201 N. Milledge Ave. Athens, GA USA 30602.

@@

\section{Main Text}\label{main-text}

@@

Using published data on excess mortality\citep{forzieri2017increasing},
we connect climate change excess mortality to life expectancy in a
mortality model. By estimating the increase in age-specific mortality
rates associated with previously published excess death estimates, we
can assess how much climate change could reduce the anticipated
longevity of the average person in twenty-eight European countries. This
approach allows us to quantify the impact of climate change on human
longevity and answer the following questions: What is the cost of
inaction on climate change on human longevity? and How does climate
change related mortality compare to other mortality vectors? Our results
can situate climate change mortality within the broader context of human
mortality and can be used to inform public health interventions to
prevent such futures.

We estimate changes in life expectancy by combining three primary
datasets: excess mortality data due to meterological hazards related to
climate change from Forzieri et al. \citep{forzieri2017increasing},
cause-of-death distributions from the Global Burden of Disease study
(GBD) \citep{GBD, wang2012age}, and life tables from the Human Mortality
Database (HMD) \citep{HMD}. Forzieri et al.
\citep{forzieri2017increasing} focused on the projected excess mortality
of heatwaves, coldwaves, wildfires, droughts, river and coastal floods,
and windstorms under the business-as-usual greenhouse gas emission
scenario (SRES A1B). They modelled long-term, small-area demographic
changes in European countries that correspond to a middle-of-the-road
socioeconomic scenario (SSP2 with medium fertility, medium mortality,
medium migration, and the Global Education Trend education scenario) and
based their projected mortality on extrapolations of an exhaustive
examination of contemporary disaster databases within spatially
rectified, high-resolution gridded datasets. While the disaster data are
not standardized across country, Forzeri et al. imputed data in
incomplete time periods and countries. This imputation could mask
spatial variability at the sub-national level, but here we use their
country-aggregated results. Their results represent business-as-usual
climate change and human development without incorporating potential
adaptation. They found a potential 150,000+ climate change related
fatalities per year by the mid 2080s with climate change contributing to
90\% of mortality rise as opposed to population changes. These data
provide the anticipated fatalities but not the effect these deaths might
have on life expectancy.

@@

We then recalculated age-specific mortality rates in five-year life
tables by adding the climate change mortality onto the underlying
mortality data and create four subsequent life tables (\(baseline\),
\(low\), \(mid\), \(high\)) using standard life table techniques
\citep{wunsch2013life} based on the published confidence intervals from
\citep{forzieri2017increasing}. We used a variant of the cause-deleted
life tables approach
\citep{brand2005approximations, beltran2008integrated} to measure the
impact of excess mortality on life expectancy at birth or \(e_0\).
Changes in life expectancy are reported based on differentials from
\(e_0\) between the 2080s and the most recent complete life table in the
HMD. Complete life tables for all countries are available in the
Supplementary Dataset and all equations are located in the Supplementary
Information.

We find that climate change could alter life expectancy by -0.23 years
(-2.5 months (-0.1 to -0.39 years)) in the average European country
(\textbf{\autoref{figure1}}). This reduction is comparable to mortality
due to influenza and pneumonia \citep{arias2013united} in the United
States. Although the average European country could see a change of
-0.23 years, several countries are likely to experience considerably
greater reductions in life expectancy. Luxembourg could see a reduction
of up to 2.2 years and the medium-variant of climate hazards could cost
the average Spaniard 1.03 years of life.

\begin{figure}
\centering
\includegraphics{MS-cclifeexpec_files/figure-latex/figure1-1.pdf}
\caption{Change in life expectancy at birth (\(e_0\)) due to
business-as-usual climate change. We report changes in life expectancies
due to climate change for twenty-eight European countries. The central
values represent the ensemble median while the stems represent the upper
and lower bounds of the inter-model climate variability.\label{figure1}}
\end{figure}

Our results also suggest climate change mortality differentials are
likely to unfold along highly uneven geographies. Whereas many Northern
European countries could experience negligible impacts on life
expectancy, five European countries could see life expectancy changes in
excess of -0.5 years (Spain, Luxembourg, France, Italy, and Ireland).
This group of countries could see life expectancy changes of more than
-1.0 years if climate change hazards are more intense than anticipated.
In some countries, climate change could thus become a bigger killer than
trachea, bronchus, and lung malignant neoplasms (-0.85 years), acute
myocardial infarction (-0.87 years), or all accident related mortality
(-0.84 years) \citep{arias2013united} by the end of the century.

@@

In this article, we demonstrate the impact climate change could have on
life expectancy at birth in twenty-eight European countries. Previous
studies on climate change and excess mortality potentially
miscommunicate the impact climate change could have on human mortality.
Contrary to excess mortality estimates, life expectancy is routinely
used as a primary metric for communicating overall health outcomes and
enjoys widespread use by major international organizations
\citep{world2015world, marmot2012building, salomon2012healthy}. Life
expectancy and its derivatives are the recommended metrics for
population health \citep{parrish2010peer}. Additionally, it connects
mortality estimates into intuitively understandable metrics, translating
global estimates of mortality into individual outcomes. Our work reveals
the extent to which climate change could reduce the average person's
longevity by the end of the century, the first study of its kind,
significantly expanding our understanding of climate change and public
health; thus linking two of the major areas in current developmental and
sustainability discussion at both national and international levels
\citep{abel2016meeting}.

We find that climate change could alter life expectancy by -0.23 years
(-2.5 months (-0.1 - -0.39 years)) in the average European country
(\textbf{Fig. 1}). Although the average European country could see a
change of -0.22 years, several countries are likely to experience
considerably greater reductions in life expectancy. Luxembourg could see
a change of up to -2.2 years and the medium-variant of climate hazards
could cost the average Spaniard 0.98 years of life.

The overall reduction in \(e_0\) is comparable to mortality due to
influenza and pneumonia \citep{arias2013united} in the United States.
\textbf{Table 1} reports the ranking of Climate Change compared to the
major causes of death (COD) in European countries using age-standardised
death rates and the standard European population
\citep{pace2013revision} for the year 2013. In three European countries
(Spain, Luxembourg, Italy), Climate Change could become a top three COD.
In Spain (high = 101.62), Luxembourg (high = 160.93), France (high =
93.45), and Italy (high = 68.21) Climate Change could become the leading
cause of death if Climate Change is more extreme than anticipated. Only
all circulatory diseases combined and all cancers combined would cause
more deaths.

Our results also suggest climate change mortality differentials are
likely to unfold along highly uneven geographies. Whereas many Northern
European countries could experience negligible impacts on life
expectancy, five European countries could see life expectancy changes in
excess of -0.5 years (Luxembourg, Spain, Italy, France, and Ireland).
This group of countries could see life expectancy changes of more than
-1.0 years if climate change hazards are more intense than anticipated.
In some countries, climate change could thus become a bigger killer than
trachea, bronchus, and lung malignant neoplasms (-0.85 years), acute
myocardial infarction (-0.87 years), or all accident related mortality
(-0.84 years) \citep{arias2013united} by the end of the century.

While Forzieri et al's \citep{forzieri2017increasing} results suggest
that the greatest climate change related mortality will unfold along a
north-south gradient, we find the greatest reduction in life expectancy
unfolds along an east-west gradient (\textbf{Fig. 2}). The most westerly
European countries tend to have the greatest reductions. Neighboring
pairs of countries at similar latitudes could experience vastly
different mortality regimes, but neighboring pairs of countries at
similar longitudes seem to experience more similar mortality regimes
from climate change. Emergent research in Atmospheric Rivers suggest
Western Europe could be more susceptible to these types of events
\citep{ramos2015daily}. Additionally, these results point to the
importance of converting excess mortality into life expectancy to
properly quantify the effects of climate change mortality.

In this article, we demonstrate the impact climate change could have on
life expectancy at birth in twenty-eight European countries. Previous
studies on climate change and excess mortality potentially misrepresent
the impact climate change could have on human mortality. Contrary to
excess mortality estimates, life expectancy is routinely used as a
primary metric for communicating overall health outcomes and enjoys
widespread use by major international organizations
\citep{world2015world, marmot2012building, salomon2012healthy}. Life
expectancy and its derivatives are the recommended metrics for
population health \citep{parrish2010peer}. Additionally, it connects
mortality estimates into intuitively understandable metrics, translating
global estimates of mortality into individual outcomes. Our work reveals
the extent to which climate change could reduce the average person's
longevity by the end of the century, the first study of its kind,
significantly expanding our understanding of climate change and public
health; thus linking two of the major areas in current developmental and
sustainability discussion at both national and international levels
\citep{abel2016meeting}.

@@

@@

\section{Methods}\label{methods}

\subsection{Data}\label{data}

The third dataset comes from the GBD \citep{GBD, wang2012age}. The GBD
is the most comprehensive worldwide dataset of epidemiological data
produced and provides cause-specific mortality by age/sex/geography/year
for 249 causes of death in 195 countries and territories. We gathered
data on age-specific deaths and mortality rates (\(m_x\)) for mortality
from environmental heat and cold exposure (cause C.2.9) for the period
2006-2015 for each country in the study. These data provide a mortality
schedule to fit the projected climate change excess mortality from
Fozieri et al \citep{forzieri2017increasing}.

\subsection{Methods}\label{methods-1}

From the GBD data, we derived variable \(t_{x,i}\) as the proportion of
deaths, \(D\), from each age group \(x\) in each country \(i\)
(\(_nt_{x,i}=D_{x,i}^{GBD}/\sum_{\alpha=0}^{80}{_nD_i^{GBD}}\)).

\begin{equation}
_n\hat{m}_{x,i,s} = (_nD_{x,i}^{HMD} + (\hat{D_i} \,\cdot\, _nt_{x,i}) ) \,/\, _nP_{x,i}^{HMD}
\end{equation}

where \(D_{x,i}\) is the number of deaths in age group \(x\) in country
\(i\) from the HMD, \(P_{x,i}\) is the population in age group \(x\)
from the HMD, \(\hat{D_x}\) is the number of deaths from Forzieri et al.
\citep{forzieri2017increasing}, and \(t_{x,i}\) is the proportion of
mortality experienced in each age group \(x\) from the GBD. Thus, the
anticipated additional mortality due to climate change is added to each
age group based on the underlying mortality cause-specific mortality
schedule observed between 2006-2015 in the GBD.

To determine the differences between life expectancy compared to the
baseline, we simply subtract \(e_{x,i,s}\) (for scenario \(LOW\),
\(MID\), and \(HIGH\)) from \(e_{x,i,BASE}\) as described by
\citep{beltran2008integrated}.
\newpage
\singlespacing 
\bibliography{LATEX/ccmort.bib}
\end{document}

>>>>

\title{Inaction on climate change could approach one year of life in some
European countries \thanks{All data and code that supports these conclusions are available as
supplementary materials.}  }

@@

\usepackage{natbib}
\bibliographystyle{apsr}
\usepackage[strings]{underscore} % protect underscores in most circumstances

@@

\@ifpackageloaded{color}{
    \PassOptionsToPackage{usenames,dvipsnames}{color}
}{%
    \usepackage[usenames,dvipsnames]{color}
}
\makeatother
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Mathew E. Hauer\textsuperscript{1}* (University of Georgia) and Alexis R. Santos\textsuperscript{2} (Pennsylvania State University)},
             pdfkeywords = {Climate change, Life expectancy, Mortality, Demography, Excess
mortality, Europe, Public health},  
            pdftitle={Inaction on climate change could approach one year of life in some
European countries},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

\noindent Climate change related estimates of excess mortality clearly demonstrate
the dramatic impact on public health and human mortality from climate
change. However, life expectancy at birth is more easily communicated
and understood by the public, and controls for different geographic,
economic, and demographic groups. No studies have comprehensively
connected excess mortality due to climate change to anticipated
reductions in life expectancy. Without properly situating the potential
loss of life within the contexts of life expectancy, we risk
misrepresenting cost of climate change on mortality. In this paper, we
convert excess mortality estimates due to increaes in extreme weather
from climate change into potential reductions of life expectancy at
birth and report the cost of climate change on human longevity. We find
climate change extremes could become the third largest life expectancy
reducer behind heart disease and cancer in some European countries.
Thus, the cost of inaction on climate change could approach one year of
life in some countries and possibly exceed that.

@@

\textsuperscript{1} Department of Sociology, Florida State University.
201 N. Milledge Ave. Athens, GA USA 30602.

@@

\hypertarget{main-text}{%
\section{Main Text}\label{main-text}}

@@

Using published data on excess mortality \citep{forzieri2017increasing},
we connect climate change excess mortality to life expectancy in a
mortality model. By estimating the increase in age-specific mortality
rates associated with previously published excess death estimates, we
can assess how much climate change could reduce the anticipated
longevity of the average person in twenty-eight European countries. This
approach allows us to quantify the impact of climate change on human
longevity and answer the following questions: What is the cost of
inaction on climate change on human longevity? and How does climate
change related mortality compare to other mortality vectors? Our results
situate climate change mortality within the broader context of human
mortality and can be used to inform public health interventions to
prevent such futures.

Our results suggest that climate change could emerge as a potent new
mortality vector in the 20th century. Mitigation strategies (ie
reductions in greenhouse gas emisions) and adaptation strategies (ie
outreach efforts, retrofitting public buildings, etc.) would help
prevent this new public health concern.

\hypertarget{background}{%
\section{Background}\label{background}}

\hypertarget{climate-change-and-human-mortality}{%
\subsection{Climate change and human
mortality}\label{climate-change-and-human-mortality}}

\hypertarget{methods-and-materials}{%
\section{Methods and Materials}\label{methods-and-materials}}

We estimate changes in life expectancy by combining three primary
datasets: excess mortality data due to meterological hazards related to
climate change from Forzieri et al. \citep{forzieri2017increasing},
cause-of-death distributions from the Global Burden of Disease study
(GBD) \citep{GBD, wang2012age}, and life tables from the Human Mortality
Database (HMD) \citep{HMD}. Forzieri et al.
\citep{forzieri2017increasing} focused on the projected excess mortality
of heatwaves, coldwaves, wildfires, droughts, river and coastal floods,
and windstorms under the business-as-usual greenhouse gas emission
scenario (SRES A1B). They modelled long-term, small-area demographic
changes in European countries that correspond to a middle-of-the-road
socioeconomic scenario (SSP2 with medium fertility, medium mortality,
medium migration, and the Global Education Trend education scenario) and
based their projected mortality on extrapolations of an exhaustive
examination of contemporary disaster databases within spatially
rectified, high-resolution gridded datasets. While the disaster data are
not standardized across country, Forzeri et al.~imputed data in
incomplete time periods and countries. This imputation could mask
spatial variability at the sub-national level, but here we use their
country-aggregated results. Their results represent business-as-usual
climate change and human development without incorporating potential
adaptation. They found a potential 150,000+ climate change related
fatalities per year by the mid 2080s with climate change contributing to
90\% of mortality rise as opposed to population changes. These data
provide the anticipated fatalities but not the effect these deaths might
have on life expectancy.

@@

\begin{table}
\caption{\textbf{Most recent mortality data available by country in the HMD.}} \label{Table2}
\begin{tabular}{ll|ll|ll|ll}
Country & Year & Country & Year & Country & Year & Country & Year \\
\hline
AUT     & 2014 & ESP     & 2014 & HUN     & 2014 & NLD     & 2016 \\
BEL     & 2015 & EST     & 2014 & IRL     & 2014 & NOR     & 2014 \\
BGR     & 2010 & FIN     & 2015 & ISL     & 2016 & POL     & 2014 \\
CHE     & 2016 & FRATNP  & 2015 & ITA     & 2014 & PRT     & 2015 \\
CZE     & 2016 & GBR     & 2016 & LTU     & 2014 & SVK     & 2014 \\
DEUT    & 2015 & GRC     & 2013 & LUX     & 2014 & SVN     & 2014 \\
DNK     & 2016 & HRV     & 2016 & LVA     & 2014 & SWE     & 2016 \\
\hline
\end{tabular}
\end{table}

We then recalculated age-specific mortality rates in five-year life
tables by adding the climate change mortality onto the underlying
mortality data and create four subsequent life tables (\(baseline\),
\(low\), \(mid\), \(high\)) using standard life table techniques
\citep{wunsch2013life} based on the published confidence intervals from
\citep{forzieri2017increasing}. We used a variant of the cause-deleted
life tables approach
\citep{brand2005approximations, beltran2008integrated} to measure the
impact of excess mortality on life expectancy at birth or \(e_0\).
Changes in life expectancy are reported based on differentials from
\(e_0\) between the 2080s and the most recent complete life table in the
HMD (see \textbf{Table 1} for the years used). Complete life tables for
all countries are available in the \emph{Supplementary Dataset} and all
computer code to replicate our results are available online\footnote{Replication
  files are located here:
  \url{https://osf.io/fp52x/?view_only=754d9a72a2ea4f6b8e0c193dc9a590d1}}.

\hypertarget{data}{%
\subsection{Data}\label{data}}

The third dataset comes from the GBD \citep{GBD, wang2012age}. The GBD
is the most comprehensive worldwide dataset of epidemiological data
produced and provides cause-specific mortality by age/sex/geography/year
for 249 causes of death in 195 countries and territories. We gathered
data on age-specific deaths and mortality rates (\(m_x\)) for mortality
from environmental heat and cold exposure (cause C.2.9) for the period
2006-2015 for each country in the study. These data provide a mortality
schedule to fit the projected climate change excess mortality from
Fozieri et al \citep{forzieri2017increasing}. Thus, we assume that the
age-specific mortality schedules observed between 2006 and 2015 due to
environmental heat and cold exposure in the GBD (see \autoref{suppfig})
are likely to remain unchanged in the future.

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}

From the GBD data, we derived variable \(t_{x,i}\) as the proportion of
deaths, \(D\), from each age group \(x\) in each country \(i\)
(\(_nt_{x,i}=D_{x,i}^{GBD}/\sum_{\alpha=0}^{80}{_nD_{\alpha,i}^{GBD}}\)).

\begin{equation}
_n\hat{m}_{x,i,s} = (_nD_{x,i}^{HMD} + (\hat{D_{i,s}} \,\cdot\, _nt_{x,i}) ) \,/\, _nP_{x,i}^{HMD}
\end{equation}

where \(D_{x,i}\) is the number of deaths in age group \(x\) in country
\(i\) from the HMD, \(P_{x,i}\) is the population in age group \(x\)
from the HMD, \(\hat{D_{i,s}}\) is the number of deaths from Forzieri et
al. \citep{forzieri2017increasing} under scenario s, and \(t_{x,i}\) is
the proportion of mortality experienced in each age group \(x\) from the
GBD. Thus, the anticipated additional mortality due to climate change is
added to each age group based on the underlying cause-specific mortality
schedule observed between 2006-2015 in the GBD.

To determine the differences between life expectancy compared to the
baseline, we simply subtract \(e_{x,i,s}\) (for scenario \(LOW\),
\(MID\), and \(HIGH\)) from \(e_{x,i,BASE}\) as described by
\citep{beltran2008integrated}.

\hypertarget{results}{%
\section{Results}\label{results}}

We find that climate change could alter life expectancy by -0.23 years
(-2.5 months (-0.1 to -0.39 years)) in the average European country
(\textbf{\autoref{figure1}}). This reduction is comparable to mortality
due to influenza and pneumonia \citep{arias2013united} in the United
States for the year 2000. Although the average European country could
see a change of -0.23 years, several countries are likely to experience
considerably greater reductions in life expectancy. Luxembourg could see
a reduction of up to 2.2 years and the medium-variant of climate hazards
could cost the average Spaniard 1.03 years of life.

\begin{figure}
\centering
\includegraphics{MS-cclifeexpec_files/figure-latex/figure1-1.pdf}
\caption{Change in life expectancy at birth (\(e_0\)) due to
business-as-usual climate change in the 2080s compared to the present
\(e_0\). We report changes in life expectancies due to climate change
for twenty-eight European countries. The central values represent the
ensemble median while the stems represent the upper and lower bounds of
the inter-model climate variability.\label{figure1}}
\end{figure}

Our results also suggest climate change mortality differentials are
likely to unfold along highly uneven geographies (\autoref{map}).
Whereas many Northern European countries could experience negligible
impacts on life expectancy, five European countries could see life
expectancy changes in excess of -0.5 years (Spain, Luxembourg, France,
Italy, and Ireland). This group of countries could see life expectancy
changes of more than -1.0 years if climate change hazards are more
intense than anticipated. In some countries, climate change could thus
become a bigger killer than trachea, bronchus, and lung malignant
neoplasms (-0.85 years), acute myocardial infarction (-0.87 years), or
all accident related mortality (-0.84 years) \citep{arias2013united} by
the end of the century.

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

In this article, we demonstrate the impact climate change could have on
life expectancy at birth in twenty-eight European countries. Previous
studies on climate change and excess mortality potentially
miscommunicate the impact climate change could have on human mortality.
Contrary to excess mortality estimates, life expectancy is routinely
used as a primary metric for communicating overall health outcomes and
enjoys widespread use by major international organizations
\citep{world2015world, marmot2012building, salomon2012healthy}. Life
expectancy and its derivatives are the recommended metrics for
population health \citep{parrish2010peer}. Additionally, it connects
mortality estimates into intuitively understandable metrics, translating
global estimates of mortality into individual outcomes. Our work reveals
the extent to which climate change could reduce the average person's
longevity by the end of the century, expanding our understanding of
climate change and public health; thus linking two of the major areas in
current developmental and sustainability discussion at both national and
international levels \citep{abel2016meeting}.

@@

These results should also be considered conservative when compared to
the broader impact of climate change on human longevity. The disaster
databases that Forzieri et al \citep{forzieri2017increasing} use in
generating their excess mortality estimates probably, though the
disaster databases are unclear, only account for the deaths certified as
\emph{directly} caused by these hazards and are unlikely to capture the
overall numbers associated with these deaths. If the certification of
deaths due to weather extremes is similar to the certication in the
present, than our results and those of Forzieri et al reflect the exess
mortality directly attributable to climatic extremes. Despite this
limitation, the impact of these extremes on \(e_0\) is considerable,
even if conservative.

@@

\newpage
\newpage
\singlespacing 
\bibliography{LATEX/ccmort.bib}
\end{document}
